The Two Dualities of Computation: Negative and Fractional Types

### 2012-02-29 rational.tex
intro

\begin{abstract}
  Every functional programmer knows about sum and product types, {{a+b}} and
  {{a*b}}. Negative and fractional types, $a-b$ and $a/b$ respectively, are
  much less known and their computational interpretation is unfamiliar and
  often complicated. We show that in a programming model in which information
  is preserved (such as the model introduced in our recent paper on
  \emph{Information Effects}), these types have particularly natural
  computational interpretations. Intuitively, values of negative types are
  values that flow ``backwards'' to satisfy demands and values of fractional
  types are values that impose constraints on their context.  The combination
  of these negative and fractional types enables greater flexibility in
  programming by breaking global invariants into local ones that can be
  autonomously satisfied by a subcomputation. Furthermore, these types and
  their properties suggest that the previously observed duality of
  computation conflated two orthogonal notions: an additive duality that
  corresponds to backtracking and a multiplicative duality that corresponds
  to unification.
\end{abstract}

@@

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclclclclcl}
a &=& a + 0 &=& a + ((-a) + a) &=& (a + (-a)) + a &=& 0 + a &=& a
\end{array}\]
Although seemingly pointless, this isomorphism has a non-trivial and
interesting computational interpretation (see Sec.~\ref{sub:neg}).

corresponds to a type isomorphism between the object $a$ and itself.

 in our
model. For example,

On the left we have a producer
that can supply a value of type $a$ and on the right we have a consumer that
demands a value of type $a$. For the purposes of the example, think of the
producer as a buyer who has \$20.00 and the consumer as a seller who expects
\$20.00 for the merchandise. In a conventional computational model, one might
realize this situation by simply writing the identity function: the buyer
hands the money to the seller to finish the transaction. The above sequence
of isomorphisms implements this identity function in a much more interesting
way, however. 
of the above series of isomorphisms is that money is ``created'' on the
consumer side together with a corresponding debt that travels ``backwards''
until it is paid for by the money at the supplier. In other words, this
computation models a credit card transaction. It shares the same benefits of
such transactions: allowing transactions to occur without necessarily
demanding that the resources be present at the time of the transaction.

Indeed, in 
the above example, the computation on the consumer side creates

\newpage

@@

>>>>

\begin{abstract}
  Every functional programmer knows about sum and product types, {{a+b}} and
  {{a*b}} respectively. Negative and fractional types, $a-b$ and $a/b$
  respectively, are much less known and their computational interpretation is
  unfamiliar and often complicated. We show that in a programming model in
  which information is preserved (such as the model introduced in our recent
  paper on \emph{Information Effects}), these types have particularly natural
  computational interpretations. Intuitively, values of negative types are
  values that flow ``backwards'' to satisfy demands and values of fractional
  types are values that impose constraints on their context.  The combination
  of these negative and fractional types enables greater flexibility in
  programming by breaking global invariants into local ones that can be
  autonomously satisfied by a subcomputation. Furthermore, these types and
  their properties suggest that the previously observed duality of
  computation conflated two orthogonal notions: an additive duality that
  corresponds to backtracking and a multiplicative duality that corresponds
  to unification.
\end{abstract}

@@

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclclclclcl}
a &=& a + 0 &=& a + ((-a) + a) \\
  &=& (a + (-a)) + a &=& 0 + a &=& a
\end{array}\]

Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism with a non-trivial and interesting computational
interpretation.  On the left we have a producer that can supply a value of
type $a$ and on the right we have a consumer that demands a value of type
$a$. For the purposes of the example, think of the producer as someone having
\$20.00 and the consumer as someone demanding \$20.00. It is certainly
possible to directly connect the producer and consumer: such a connection
would be modeled by the identity function. The proof above implements this
connection in a much more interesting way, however. A transformation $0 =
(-a) + a$ is used at the consumer site to create money and a corresponding
debt. The consumer gets the money immediately and the debt travels
``backwards'' towards the producer where it is paid.  In other words, this
computation models something similar to a credit card transaction. We discuss
this example in detail in Sec.~\ref{sub:neg} and only focus on two important
points here:
\begin{itemize}
\item As the example illustrates, negative types corresponds to ``debts.''
  It would clearly be disastrous if debts could be deleted or
  duplicated. This simple observation explains why negative types are much
  simpler and much more appealing in a framework where information is
  guaranteed to be preserved. In previous work that used negative types,
  complicated mechanisms are typically needed to constrain the propagation
  and use of negative values because the surrounding computational framework
  is, generally speaking, careless in its treatment of information.
\item The appeal of negative types can already be gleaned from the credit
  card analogy. The main reason credit card transactions are convenient is
  because they disentangle the propagation of the money from the propagation
  of the services. Not every transaction needs both the resources and
  services to be brought together: it is sufficient to have a promise that
  the money will be paid at some point, as long as the infrastructure can be
  trusted that such promises will be fulfilled. 
\end{itemize}

Similar observations hold for fractional types but we defer these to
Sec.~\ref{sub:frac}. Before proceeding to the main part of the paper, we
briefly summarize the our technical contributions:
\begin{itemize}
\item 
\item 
\item 
\item 
\end{itemize}

@@

The simplest way to connect the In a conventional computational model, one
might realize this situation by simply writing the identity function: the
buyer hands the money to the seller to finish the transaction. The above
sequence of isomorphisms implements this identity function in a much more
interesting way, however.  of the above series of
